### 2024-07-16 journal/algorithm-GC.tex
some final edits from proofreading
cleaning tex files (removing comments etc)

The \fld{blocks} field of each node in the \ordering\ tree is implemented as a pointer to the root of a RBT of \typ{Blocks} rather than an infinite array.  
Each RBT is initialized with an empty block with index~0.
Any access to an entry of the \fld{blocks} array is replaced by a search in the RBT.
The node's \head\ field, which previously gave the next position to insert into the \fld{blocks} array is no
longer needed; we can instead simply find the maximum \fld{index} of any block in the RBT.
To facilitate this, \op{MaxBlock} is a query operation on the RBT that 
returns the block with the maximum \fld{index}.
We can store, in the root of the RBT, a pointer to the maximum block so that \op{MaxBlock}
can be done in \rev{$O(1)$ steps}, without affecting the step complexity of other RBT operations.
Similarly, a \op{MinBlock} query finds the block with the minimum \fld{index} in a RBT.

@@

    \gc{\State \var{h} \assign\ \Call{MaxBlock}{\var{leaf}[\var{id}].\fld{blocks}}.\fld{index}+1}
	\State \hangbox{let \var{B} be a new \typ{Block} with fields \fld{element} \assign\ \var{e},
		\fld{sum\sub{enq}}\ \assign\ $\var{leaf}[\var{id}].\fld{blocks}[\gc{h-1}].\fld{sum\sub{enq}}+1,$\\
		\fld{sum\sub{deq}}\ \assign\ $\var{leaf}[\var{id}].\fld{blocks}[\gc{h-1}].\fld{sum\sub{deq}}$,
		\gc{\fld{index}\ \assign $h$}}\label{enqNewGC}
    \State \Call{Append}{\var{B}}\Comment{\rev{append \op{Enqueue} to \var{leaf}[\var{id}].\fld{blocks}, propagate to root}}
\EndFunction{Enqueue}

@@

\Function{Object}{Dequeue}{} 
    \State \gc{\var{h} \assign\ \Call{MaxBlock}{\var{leaf}[\var{id}].\fld{blocks}}.\fld{index}+1}
    \State \hangbox{let \var{B} be a new \typ{Block} with fields \fld{element} \assign\ \nl,
	    \fld{sum\sub{enq}}\ \assign\ $\var{leaf}[\var{id}].\fld{blocks}[\gc{h-1}].\fld{sum\sub{enq}},$\\
	    \fld{sum\sub{deq}}\ \assign\ $\var{leaf}[\var{id}].\fld{blocks}[\gc{h-1}].\fld{sum\sub{deq}}+1$,
	    \gc{\fld{index} \assign $h$}}\label{deqNewGC}
    \State \Call{Append}{\var{B}}\Comment{\rev{append \op{Dequeue} to \var{leaf}[\var{id}].\fld{blocks}, propagate to root}}%
    \gc{\State \Return \Call{CompleteDeq}{\var{leaf}[\var{id}], $h$}}\Comment{\rev{compute response \op{Dequeue} should return}}
\EndFunction{Dequeue}

@@

\gc{
\Function{Object}{CompleteDeq}{\typ{Node} \var{leaf}, \typ{int} $h$} \Comment{finish propagated \dequeue\ in $\var{leaf}.\fld{blocks}[h]$}
    \State $\langle \var{b}, \var{i}\rangle$ \assign\ $\Call{IndexDequeue}{\var{leaf}, h, 1}$\Comment{\rev{find where \op{Dequeue} is in \var{root}.\fld{blocks}}}
    \State \Return $\Call{FindResponse}{b, i}$\Comment{\rev{find response \op{Dequeue} should return}}
\EndFunction{CompleteDeq}
}

@@

\Function{void}{Propagate}{\typ{Node} \var{v}} \Comment{propagate blocks from \var{v}'s children to root}
    \If{{\keywordfont not} \Call{Refresh}{\var{v}}} \label{firstRefreshGC}  \hfill \Comment{double refresh: \rev{if first fails, try again}}
        \State \Call{Refresh}{\var{v}} \label{secondRefreshGC}
    \EndIf
    \If{$\var{v} \neq \var{root}$} \hfill \Comment{recurse up tree}
        \State \Call{Propagate}{\var{v}.\fld{parent}}
    \EndIf
\EndFunction{Propagate}

@@

\Function{boolean}{Refresh}{\typ{Node} \var{v}} \Comment{try to append a new block $B$ to $\var{v}.\fld{blocks}$}%
	\gc{
	\State $T$ \assign\ $\var{v}.\fld{blocks}$\label{refreshReadTGC}
    \State $\var{h}$ \assign\ $\Call{MaxBlock}{T}.\fld{index}+1$ \label{refreshReadMax}
    }
    \State $\var{B}$ \assign\ $\Call{CreateBlock}{\var{v, h}}$\Comment{\rev{create new block to append to \var{v}.\fld{blocks}}}\label{invokeCreateBlockGC}
    \If{\var{B = \nl}} \Return{\tr}\Comment{\rev{if no new operations to propagate to \var{v}, do nothing}}\label{addOPGC}
    \Else  
    	\gc{
	    \State $T'$ \assign\ $\Call{AddBlock}{v, T, \var{B}}$\Comment{\rev{$T'$ is a copy of $T$ with $B$ added (and GC if needed)}}\label{refreshABGC}
		\State \Return \Call{CAS}{\var{v}.\fld{blocks}, $T$, $T'$}\Comment{\rev{try installing $T'$ as \var{v}'s \fld{blocks} tree}}\label{casGC}%
		}
    \EndIf
\EndFunction{Refresh}

@@

\gc{
\Function{RBT}{AddBlock}{\typ{Node} \var{v}, \typ{RBT} \var{T}, \typ{Block} \var{B}} \Comment{do GC if necessary; add block $\var{B}\neq\nl$ to \var{T}}
    \If{\var{B}.\fld{index} is a multiple of $G$} \Comment{do garbage collection}\label{GCstart}
        \State $s$ \assign\ \Call{SplitBlock}{\var{v}}.\var{index}\Comment{\rev{find lowest index of block to keep}}\label{callSplitBlock}
        \State \Call{Help}{}\Comment{\rev{help complete pending \op{Dequeues}}}\label{invokeHelp}
        \State $T'$ \assign\ \Call{Split}{$T,s$} \Comment{\op{Split} removes blocks with $\fld{index} < s$}\label{splitGC}
        \State \Return \Call{Insert}{$T',B$}\Comment{\rev{add $B$ to the tree}}\label{GCend}
	\Else\ \Return \Call{Insert}{$T, B$}\Comment{\rev{if GC not needed, just add $B$ to the tree}}\label{noGC-insert}
    \EndIf
\EndFunction{AddBlock}
}

@@

\begin{figure*}
\begin{algorithmic}[1]
\setcounter{ALG@line}{251}
\gc{
\Function{Block}{SplitBlock}{\typ{Node} v} \Comment{\rev{find lowest index of block GC should keep in \var{v}}}
	\If{$\var{v}=\var{root}$} \Comment{return block before the last block containing a}
		\State $m$ \assign\ 0\Comment{dequeued \enqueue\ or null \dequeue}
		\For{$k$ \assign\ $1..p$} $m$ \assign\ $\max(m, \var{v}.\fld{last}[k])$
		\EndFor
		\State $B$ \assign\ $\var{root}.\fld{blocks}[m-1]$\label{SBlookup1}
	\Else \Comment{\rev{return last sublock of first block kept in parent}}
		\State $B_p$ \assign\ $\Call{SplitBlock}{\var{v}.\fld{parent}}$\label{SBrecurse}
		\State \var{dir} \assign\ ($\var{v}=\var{v}.\fld{parent.left}$ ? \fld{left} : \fld{right})
		\State $B$ \assign\ $\var{v}.\fld{blocks}[B_p.\edir]$\label{SBlookup2}
	\EndIf
	\State \Return ($B = \nl$ ? $\Call{MinBlock}{\var{v}.\var{blocks}}$ : $B$) \Comment{If $B$ was discarded, keep all blocks}\label{substitute}
\EndFunction{SplitBlock}
}

@@

    \For{$\ell$ in $\var{leaf}[1..k]$}
        \State $\var{B}$ \assign\ $\Call{MaxBlock}{\ell.\fld{blocks}}$
    	%\State \Call{Propagate}{$\ell.\fld{parent}$}   <--- don't do this because recursive helping could be infinite loop
        \If{$\var{B}.\fld{element} = \nl$ and $\var{B}.\fld{index}>0$ and \Call{Propagated}{$\ell, \var{B}.\fld{index}$}} 
            \State \linecomment{operation is a propagated \dequeue}
            \State $\var{B}.\fld{response}$ \assign\ $\Call{CompleteDeq}{\ell, \var{B}.\fld{index}}$ \Comment{\rev{compute response and store it in leaf block}}
        \EndIf
    \EndFor
\EndFunction{Help}
}
\spac

\gc{
\Function{boolean}{Propagated}{\typ{Node} \var{v}, \typ{int} \var{b}}  \Comment{check if $\var{v}.\fld{blocks}[b]$ has  propagated to \var{root}}
    \State \linecomment{Precondition:  $\var{v}.\fld{blocks}[b]$ exists}
    \If {$\var{v} = \var{root}$} \Return{true}
    \Else
    	\State $T$ \assign\ $\var{v}.\fld{parent.blocks}$
		\State \fld{dir} \assign\ (\var{v}.\fld{parent.left} = \var{v} ? \fld{left} : \fld{right}) 
		\If{$\Call{MaxBlock}{T}.\edir < b$} \Return \fa\ \Comment{\rev{\var{v}.\fld{blocks}[\var{b}] has not propagated to \var{v}.\fld{parent}}}
		\Else \Comment{\rev{check if superblock of \var{v}.\fld{blocks}[\var{b}] propagated to root}}
			\State $B_p$ \assign\ minimum index block in $T$ with $\edir \geq b$\label{searchsuper3}
			\State \Return \Call{Propagated}{\var{v}.\fld{parent}, $B_p$.\fld{index}}
		\EndIf
	\EndIf
\EndFunction{Propagated}
}

@@

\begin{figure*}
\begin{algorithmic}[1]
\setcounter{ALG@line}{296}
\Function{$\langle\typ{int}, \typ{int}\rangle$}{IndexDequeue}{\typ{Node} \var{v}, \typ{int} \var{b}, \typ{int} \var{i}}
    \State \linecomment{return $\langle b', i' \rangle$ such that \var{i}th \dequeue\ in $D(\var{v}.\fld{blocks}[\var{b}])$ is ($i'$)th \dequeue\ of $D(\var{root}.\fld{blocks}[b'])$}
    \State \linecomment{Precondition: \var{v}.\fld{blocks}[\var{b}] exists and has propagated to root and $|D(\var{v}.\fld{blocks}[\var{b}])|\geq \var{i}$}
    \If{$\var{v} = \var{root}$} \Return $\langle\var{b, i}\rangle$ \label{indexBaseCaseGC}
    \Else \Comment{First, find the superblock $B_p$ of $v.blocks[b]$}
	    \State \fld{dir} \assign\ (\var{v}.\fld{parent.left} = \var{v} ? \fld{left} : \fld{right}) 
    	\gc{
		\State $T$ \assign\ $\var{v}.\fld{parent.blocks}$
		\State $B_p$ \assign\ min block in $T$ with $\edir \geq b$\label{searchsuper1}
		\State $B_p'$ \assign\ max block in $T$ with $\edir < b$ \Comment{predecessor of $B_p$}\label{searchsuper2}}
	    \State \linecomment{compute index \var{i} of \dequeue\ in superblock $B_p$}
	    \State \linecomment{\rev{add \# \dequeues\ in blocks of $v$ before $v.\fld{blocks}[b]$ within superblock $B_p$}}
	    \State \hangbox{\var{i} += $\var{v}.\fld{blocks}[\var{b}-1].\fld{sum\sub{deq}} -
	    		\var{v}.\fld{blocks}[\gc{B_p'}.\edir].\fld{sum\sub{deq}}$}
        \If{$\fld{dir} = \fld{right}$}\Comment{\rev{add \# \dequeues\ in $B_p$ from left sibling}}
        	\State \hangbox{\var{i} += $\var{v}.\fld{blocks}[\gc{B_p}.\eleft].\fld{sum\sub{deq}} - \mbox{ }
					\var{v}.\fld{blocks}[\gc{B_p'}.\eleft].\fld{sum\sub{deq}}$}\label{considerLeftBeforeRightGC}
        \EndIf \label{computeISuperEndGC}
        \State \Return\Call{IndexDequeue}{\var{v}.\fld{parent}, $\gc{B_p.\fld{index}}$, \var{i}} \Comment{\rev{recurse up the tree}}
    \EndIf
\EndFunction{IndexDequeue}

@@

		\State use \gc{BST search in \var{root}.\fld{blocks} to find minimum index $b_e$ of a \typ{Block} with $\fld{sum\sub{enq}} \geq \var{e}$}\label{FRsearchGC} 
		\State $i_e$ \assign\ $\var{e} - \var{root}.\fld{blocks}[b_e-1].\fld{sum\sub{enq}}$ \Comment{find rank of \enqueue\ within its block}%
		\gc{
        \State \var{res} \assign\ \Call{GetEnqueue}{\var{root}, $b_e$, $i_e$}\Comment{\rev{find argument of that \enqueue}}\label{findAnswerGC}
        \If {$b_e > \var{last}[\var{id}]$} $\var{last}[\var{id}]$ \assign\ $b_e$
        \EndIf
        \State \Return \var{res}
		}
    \EndIf
\EndFunction{FindResponse}

@@

\Function{element}{GetEnqueue}{\typ{Node} \var{v}, \typ{int} \var{b}, \typ{int} \var{i}} \Comment{returns argument of \var{i}th \enqueue\ in $E(\var{v}.\fld{blocks}[\var{b}])$}
    \State \linecomment{Preconditions: $\var{i}\geq 1$ and \var{v}.\fld{blocks}[\var{b}] exists and contains at least \var{i} \enqueues}
    \If{\var{v} is a leaf node} \Return \var{v}.\fld{blocks}[\var{b}].\fld{element} \label{getBaseCaseGC}
    \Else 
        \State \var{sum\sub{left}} \assign\ \var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[\var{b}].\eleft].\fld{sum\sub{enq}} \Comment{\#\ \enqueues\ in \var{v}.\fld{blocks}[1..$\var{b}$] from \var{v}.\fld{left}}
        \State \var{prev\sub{left}} \assign\ \var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[$\var{b}-1$].\eleft].\fld{sum\sub{enq}} \Comment{\#\ \enqueues\ in \var{v}.\fld{blocks}[1..$\var{b}-1$] from \var{v}.\fld{left}}
        \State \var{prev\sub{right}} \assign\ \var{v}.\fld{right.blocks}[\var{v}.\fld{blocks}[$\var{b}-1$].\eright].\fld{sum\sub{enq}} \Comment{\#\ \enqueues\ in \var{v}.\fld{blocks}[1..$\var{b}-1$] from \var{v}.\fld{right}}
        \If{$\var{i} \leq \var{sum\sub{left}} - \var{prev\sub{left}}$} \fld{dir} \assign\ \fld{left}\Comment{required \enqueue\ is in \var{v}.\fld{left}}\label{leftOrRightGC} 
        \Else \Comment{required \enqueue\ is in \var{v}.\fld{right}}
            \State \fld{dir} \assign\ \fld{right}
            \State $\var{i}\ \assign\ \var{i} - (\var{sum\sub{left}} - \var{prev\sub{left}})$\Comment{\rev{deduct \# \enqueues\ in $v.\fld{blocks}[b]$ that came from $v.\fld{left}$}}
        \EndIf
        \State \linecomment{find \enqueue's block in \var{v}.\fld{dir.blocks} and its rank within block}
        \State \gc{use BST search in \var{v}.\fld{dir.blocks} to find minimum index $\var{b}'$ of a \typ{Block} with $\fld{sum\sub{enq}} \geq \var{i} + \var{prev\sub{dir}}$}\label{getChildGC}
        \State $\var{i}'$ \assign\ $\var{i} - (\var{v}.\fld{dir.blocks}[\var{b}'-1].\fld{sum\sub{enq}} - \var{prev\sub{dir}})$
        \State \Return\Call{GetEnqueue}{\var{v}.\fld{dir}, $\var{b}'$, $\var{i}'$}
    \EndIf
\EndFunction{GetEnqueue}

>>>>

The \fld{blocks} field of each node in the \ordering\ tree is implemented as a pointer to the root of a RBT of \typ{Blocks} rather than an infinite array.  
Each RBT is initialized with an empty block with index~0.
Any access to an entry of the \fld{blocks} array is replaced by a search in the RBT.
The node's \head\ field, which previously gave the next position to insert into the \fld{blocks} array is no
longer needed; we can instead simply find the maximum \fld{index} of any block in the RBT.
To facilitate this, \op{MaxBlock} is a query operation on the RBT that 
returns the block with the maximum \fld{index}.
We can store, in the root of the RBT, a pointer to the maximum block so that \op{MaxBlock}
can be done in $O(1)$ steps, without affecting the step complexity of other RBT operations.
Similarly, a \op{MinBlock} query finds the block with the minimum \fld{index} in a RBT.

@@

    \gc{\State \var{h} \assign\ \Call{MaxBlock}{\var{leaf}[\var{id}].\fld{blocks}}.\fld{index}+1}
	\State \hangbox{let \var{B} be a new \typ{Block} with fields \fld{element} \assign\ \var{e},
		\fld{sum\sub{enq}}\ \assign\ $\var{leaf}[\var{id}].\fld{blocks}[\gc{h-1}].\fld{sum\sub{enq}}+1,$\\
		\fld{sum\sub{deq}}\ \assign\ $\var{leaf}[\var{id}].\fld{blocks}[\gc{h-1}].\fld{sum\sub{deq}}$,
		\gc{\fld{index}\ \assign $h$}}\label{enqNewGC}
    \State \Call{Append}{\var{B}}\Comment{append \op{Enqueue} to \var{leaf}[\var{id}].\fld{blocks}, propagate to root}
\EndFunction{Enqueue}

@@

\Function{Object}{Dequeue}{} 
    \State \gc{\var{h} \assign\ \Call{MaxBlock}{\var{leaf}[\var{id}].\fld{blocks}}.\fld{index}+1}
    \State \hangbox{let \var{B} be a new \typ{Block} with fields \fld{element} \assign\ \nl,
	    \fld{sum\sub{enq}}\ \assign\ $\var{leaf}[\var{id}].\fld{blocks}[\gc{h-1}].\fld{sum\sub{enq}},$\\
	    \fld{sum\sub{deq}}\ \assign\ $\var{leaf}[\var{id}].\fld{blocks}[\gc{h-1}].\fld{sum\sub{deq}}+1$,
	    \gc{\fld{index} \assign $h$}}\label{deqNewGC}
    \State \Call{Append}{\var{B}}\Comment{append \op{Dequeue} to \var{leaf}[\var{id}].\fld{blocks}, propagate to root}%
    \gc{\State \Return \Call{CompleteDeq}{\var{leaf}[\var{id}], $h$}}\Comment{compute response \op{Dequeue} should return}
\EndFunction{Dequeue}

@@

\gc{
\Function{Object}{CompleteDeq}{\typ{Node} \var{leaf}, \typ{int} $h$} \Comment{finish propagated \dequeue\ in $\var{leaf}.\fld{blocks}[h]$}
    \State $\langle \var{b}, \var{i}\rangle$ \assign\ $\Call{IndexDequeue}{\var{leaf}, h, 1}$\Comment{find where \op{Dequeue} is in \var{root}.\fld{blocks}}
    \State \Return $\Call{FindResponse}{b, i}$\Comment{find response \op{Dequeue} should return}
\EndFunction{CompleteDeq}
}

@@

\Function{void}{Propagate}{\typ{Node} \var{v}} \Comment{propagate blocks from \var{v}'s children to root}
    \If{{\keywordfont not} \Call{Refresh}{\var{v}}} \label{firstRefreshGC}  \hfill \Comment{double refresh: if first \op{Refresh} fails, try again}
        \State \Call{Refresh}{\var{v}} \label{secondRefreshGC}
    \EndIf
    \If{$\var{v} \neq \var{root}$} \hfill \Comment{recurse up tree}
        \State \Call{Propagate}{\var{v}.\fld{parent}}
    \EndIf
\EndFunction{Propagate}

@@

\Function{boolean}{Refresh}{\typ{Node} \var{v}} \Comment{try to append a new block $B$ to $\var{v}.\fld{blocks}$}%
	\gc{
	\State $T$ \assign\ $\var{v}.\fld{blocks}$\label{refreshReadTGC}
    \State $\var{h}$ \assign\ $\Call{MaxBlock}{T}.\fld{index}+1$ \label{refreshReadMax}
    }
    \State $\var{B}$ \assign\ $\Call{CreateBlock}{\var{v, h}}$\Comment{create new block to append to \var{v}.\fld{blocks}}\label{invokeCreateBlockGC}
    \If{\var{B = \nl}} \Return{\tr}\Comment{if no new operations to propagate to \var{v}, do nothing}\label{addOPGC}
    \Else  
    	\gc{
	    \State $T'$ \assign\ $\Call{AddBlock}{v, T, \var{B}}$\Comment{$T'$ is a copy of $T$ with $B$ added (and GC if needed)}\label{refreshABGC}
		\State \Return \Call{CAS}{\var{v}.\fld{blocks}, $T$, $T'$}\Comment{try installing $T'$ as \var{v}'s \fld{blocks} tree}\label{casGC}%
		}
    \EndIf
\EndFunction{Refresh}

@@

\gc{
\Function{RBT}{AddBlock}{\typ{Node} \var{v}, \typ{RBT} \var{T}, \typ{Block} \var{B}} \Comment{do GC if necessary; add block $\var{B}\neq\nl$ to \var{T}}
    \If{\var{B}.\fld{index} is a multiple of $G$} \Comment{do garbage collection}\label{GCstart}
        \State $s$ \assign\ \Call{SplitBlock}{\var{v}}.\var{index}\Comment{find lowest index of block to keep}\label{callSplitBlock}
        \State \Call{Help}{}\Comment{help complete pending \op{Dequeues}}\label{invokeHelp}
        \State $T'$ \assign\ \Call{Split}{$T,s$} \Comment{\op{Split} removes blocks with $\fld{index} < s$}\label{splitGC}
        \State \Return \Call{Insert}{$T',B$}\Comment{add $B$ to the tree}\label{GCend}
	\Else\ \Return \Call{Insert}{$T, B$}\Comment{if GC not needed, just add $B$ to the tree}\label{noGC-insert}
    \EndIf
\EndFunction{AddBlock}
}

@@

\begin{figure*}
\begin{algorithmic}[1]
\setcounter{ALG@line}{251}
\gc{
\Function{Block}{SplitBlock}{\typ{Node} v} \Comment{find lowest index of block GC should keep in \var{v}}
	\If{$\var{v}=\var{root}$} \Comment{return block before the last block containing a}
		\State $m$ \assign\ 0\Comment{dequeued \enqueue\ or null \dequeue}
		\For{$k$ \assign\ $1..p$} $m$ \assign\ $\max(m, \var{v}.\fld{last}[k])$
		\EndFor
		\State $B$ \assign\ $\var{root}.\fld{blocks}[m-1]$\label{SBlookup1}
	\Else \Comment{return last sublock of first block kept in parent}
		\State $B_p$ \assign\ $\Call{SplitBlock}{\var{v}.\fld{parent}}$\label{SBrecurse}
		\State \var{dir} \assign\ ($\var{v}=\var{v}.\fld{parent.left}$ ? \fld{left} : \fld{right})
		\State $B$ \assign\ $\var{v}.\fld{blocks}[B_p.\edir]$\label{SBlookup2}
	\EndIf
	\State \Return ($B = \nl$ ? $\Call{MinBlock}{\var{v}.\var{blocks}}$ : $B$) \Comment{If $B$ was discarded, keep all blocks}\label{substitute}
\EndFunction{SplitBlock}
}

@@

    \For{$\ell$ in $\var{leaf}[1..k]$}
        \State $\var{B}$ \assign\ $\Call{MaxBlock}{\ell.\fld{blocks}}$
    	%\State \Call{Propagate}{$\ell.\fld{parent}$}   <--- don't do this because recursive helping could be infinite loop
        \If{$\var{B}.\fld{element} = \nl$ and $\var{B}.\fld{index}>0$ and \Call{Propagated}{$\ell, \var{B}.\fld{index}$}} 
            \State \linecomment{operation is a propagated \dequeue}
            \State $\var{B}.\fld{response}$ \assign\ $\Call{CompleteDeq}{\ell, \var{B}.\fld{index}}$ \Comment{compute response and store it in leaf block}
        \EndIf
    \EndFor
\EndFunction{Help}
}
\spac

\gc{
\Function{boolean}{Propagated}{\typ{Node} \var{v}, \typ{int} \var{b}}  \Comment{check if $\var{v}.\fld{blocks}[b]$ has  propagated to \var{root}}
    \State \linecomment{Precondition:  $\var{v}.\fld{blocks}[b]$ exists}
    \If {$\var{v} = \var{root}$} \Return{true}
    \Else
    	\State $T$ \assign\ $\var{v}.\fld{parent.blocks}$
		\State \fld{dir} \assign\ (\var{v}.\fld{parent.left} = \var{v} ? \fld{left} : \fld{right}) 
		\If{$\Call{MaxBlock}{T}.\edir < b$} \Return \fa\ \Comment{\var{v}.\fld{blocks}[\var{b}] has not propagated to \var{v}.\fld{parent}}
		\Else \Comment{check if superblock of \var{v}.\fld{blocks}[\var{b}] propagated to root}
			\State $B_p$ \assign\ minimum index block in $T$ with $\edir \geq b$\label{searchsuper3}
			\State \Return \Call{Propagated}{\var{v}.\fld{parent}, $B_p$.\fld{index}}
		\EndIf
	\EndIf
\EndFunction{Propagated}
}

@@

\begin{figure*}
\begin{algorithmic}[1]
\setcounter{ALG@line}{296}
\Function{$\langle\typ{int}, \typ{int}\rangle$}{IndexDequeue}{\typ{Node} \var{v}, \typ{int} \var{b}, \typ{int} \var{i}}
    \State \linecomment{return $\langle b', i' \rangle$ such that \var{i}th \dequeue\ in $D(\var{v}.\fld{blocks}[\var{b}])$ is ($i'$)th \dequeue\ of $D(\var{root}.\fld{blocks}[b'])$}
    \State \linecomment{Precondition: \var{v}.\fld{blocks}[\var{b}] exists and has propagated to root and $|D(\var{v}.\fld{blocks}[\var{b}])|\geq \var{i}$}
    \If{$\var{v} = \var{root}$} \Return $\langle\var{b, i}\rangle$ \label{indexBaseCaseGC}
    \Else \Comment{First, find the superblock $B_p$ of $v.blocks[b]$}
	    \State \fld{dir} \assign\ (\var{v}.\fld{parent.left} = \var{v} ? \fld{left} : \fld{right}) 
    	\gc{
		\State $T$ \assign\ $\var{v}.\fld{parent.blocks}$
		\State $B_p$ \assign\ min block in $T$ with $\edir \geq b$\label{searchsuper1}
		\State $B_p'$ \assign\ max block in $T$ with $\edir < b$ \Comment{predecessor of $B_p$}\label{searchsuper2}}
	    \State \linecomment{compute index \var{i} of \dequeue\ in superblock $B_p$}
	    \State \linecomment{add \# \dequeues\ in blocks of $v$ before $v.\fld{blocks}[b]$ within superblock $B_p$}
	    \State \hangbox{\var{i} += $\var{v}.\fld{blocks}[\var{b}-1].\fld{sum\sub{deq}} -
	    		\var{v}.\fld{blocks}[\gc{B_p'}.\edir].\fld{sum\sub{deq}}$}
        \If{$\fld{dir} = \fld{right}$}\Comment{add \# \dequeues\ in $B_p$ from left sibling}
        	\State \hangbox{\var{i} += $\var{v}.\fld{blocks}[\gc{B_p}.\eleft].\fld{sum\sub{deq}} - \mbox{ }
					\var{v}.\fld{blocks}[\gc{B_p'}.\eleft].\fld{sum\sub{deq}}$}\label{considerLeftBeforeRightGC}
        \EndIf \label{computeISuperEndGC}
        \State \Return\Call{IndexDequeue}{\var{v}.\fld{parent}, $\gc{B_p.\fld{index}}$, \var{i}} \Comment{recurse up the tree}
    \EndIf
\EndFunction{IndexDequeue}

@@

		\State use \gc{BST search in \var{root}.\fld{blocks} to find minimum index $b_e$ of a \typ{Block} with $\fld{sum\sub{enq}} \geq \var{e}$}\label{FRsearchGC} 
		\State $i_e$ \assign\ $\var{e} - \var{root}.\fld{blocks}[b_e-1].\fld{sum\sub{enq}}$ \Comment{find rank of \enqueue\ within its block}%
		\gc{
        \State \var{res} \assign\ \Call{GetEnqueue}{\var{root}, $b_e$, $i_e$}\Comment{find argument of that \enqueue}\label{findAnswerGC}
        \If {$b_e > \var{last}[\var{id}]$} $\var{last}[\var{id}]$ \assign\ $b_e$
        \EndIf
        \State \Return \var{res}
		}
    \EndIf
\EndFunction{FindResponse}

@@

\Function{element}{GetEnqueue}{\typ{Node} \var{v}, \typ{int} \var{b}, \typ{int} \var{i}} \Comment{returns argument of \var{i}th \enqueue\ in $E(\var{v}.\fld{blocks}[\var{b}])$}
    \State \linecomment{Preconditions: $\var{i}\geq 1$ and \var{v}.\fld{blocks}[\var{b}] exists and contains at least \var{i} \enqueues}
    \If{\var{v} is a leaf node} \Return \var{v}.\fld{blocks}[\var{b}].\fld{element} \label{getBaseCaseGC}
    \Else 
        \State \var{sum\sub{left}} \assign\ \var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[\var{b}].\eleft].\fld{sum\sub{enq}} \Comment{\#\ \enqueues\ in \var{v}.\fld{blocks}[1..$\var{b}$] from \var{v}.\fld{left}}
        \State \var{prev\sub{left}} \assign\ \var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[$\var{b}-1$].\eleft].\fld{sum\sub{enq}} \Comment{\#\ \enqueues\ in \var{v}.\fld{blocks}[1..$\var{b}-1$] from \var{v}.\fld{left}}
        \State \var{prev\sub{right}} \assign\ \var{v}.\fld{right.blocks}[\var{v}.\fld{blocks}[$\var{b}-1$].\eright].\fld{sum\sub{enq}} \Comment{\#\ \enqueues\ in \var{v}.\fld{blocks}[1..$\var{b}-1$] from \var{v}.\fld{right}}
        \If{$\var{i} \leq \var{sum\sub{left}} - \var{prev\sub{left}}$} \fld{dir} \assign\ \fld{left}\Comment{required \enqueue\ is in \var{v}.\fld{left}}\label{leftOrRightGC} 
        \Else \Comment{required \enqueue\ is in \var{v}.\fld{right}}
            \State \fld{dir} \assign\ \fld{right}
            \State $\var{i}\ \assign\ \var{i} - (\var{sum\sub{left}} - \var{prev\sub{left}})$\Comment{deduct \# \enqueues\ in $v.\fld{blocks}[b]$ that came from $v.\fld{left}$}
        \EndIf
        \State \linecomment{find \enqueue's block in \var{v}.\fld{dir.blocks} and its rank within block}
        \State \gc{use BST search in \var{v}.\fld{dir.blocks} to find minimum index $\var{b}'$ of a \typ{Block} with $\fld{sum\sub{enq}} \geq \var{i} + \var{prev\sub{dir}}$}\label{getChildGC}
        \State $\var{i}'$ \assign\ $\var{i} - (\var{v}.\fld{dir.blocks}[\var{b}'-1].\fld{sum\sub{enq}} - \var{prev\sub{dir}})$
        \State \Return\Call{GetEnqueue}{\var{v}.\fld{dir}, $\var{b}'$, $\var{i}'$}
    \EndIf
\EndFunction{GetEnqueue}
